A Guide to Designing Experiments to Test Statistical Graphics

### 2022-07-05 index.qmd
tightening up wording

The experiment was a full 2x3x3 factorial exploration of three data generating parameters, with 3 replicates at each parameter combination (54 datasets) and 10 aesthetic combinations (for a total of 540 lineups). Each lineup had 20 different sub-panels, so we should have carefully visually inspected some 10800 different panels. 
As is evident from the fact that we're telling this story as a cautionary tale, we missed a critical problem with our data-generating mechanism: when clusters were assigned to randomly generated data after the fact, we didn't control the cluster size, leading to clusters of one or two points in relatively few sub-panels. 
This became particularly noticeable when bounding ellipses were added to the plot, as the method used to generate those ellipses required at least 3 points in the cluster. 
Many sub-plots were generated with only 2 or 4 ellipses (with 3 or 5 total clusters), which escaped our notice during the stimuli proof-reading phase of the experiment, but did not escape the notice of our participants, who only needed to examine about 10 lineups each (around 200 panels). 
An example of one of the problematic lineups is shown in @fig-lineup-problems: many participants selected panel 16 because of the missing ellipse; not a wrong choice, but certainly not the visual effect we intended to test.

@@

One reason why it is so difficult to generate sampling models for visual explorations is that our visual systems are very, very good at finding differences and unexpected results. 
We re-ran the experiment using a different clustering method, and found that instead of cuing in on the number of ellipses, when that variable was removed participants instead cued in on the differences in size and shape of the ellipses formed using k-means clustering after the data generating procedure. 
That is, participants could still detect the artificial nature of the induced clusters using other features. 
While it can be difficult to get the data generating method right, it is essential to conducting visual experiments that generalize well beyond the effects shown in a single data set or phenomenon. 
The time and effort invested in this step at the outset of the experiment pays dividends when it allows for clear generalization of the experimental results to an entire statistical concept rather than a single data set.

>>>>

The experiment was a full 2x3x3 factorial exploration of three data generating parameters, with 3 replicates at each parameter combination (54 datasets) and 10 aesthetic combinations (for a total of 540 lineups). Each lineup had 20 different sub-panels, so we should have carefully visually inspected some 10800 different panels. 
As is evident from the fact that we're telling this story as a cautionary tale, we missed a critical problem with our data-generating mechanism: when clusters were assigned to randomly generated data after the fact, we didn't control the cluster size, leading to clusters of one or two points in relatively few sub-panels. 
This became particularly noticeable when bounding ellipses were added to the plot, as the method used to generate those ellipses required at least 3 points in the cluster. 
Many sub-plots were generated with only 2 or 4 ellipses (with 3 or 5 total clusters), which escaped our notice during the stimuli proof-reading phase of the experiment, but did not escape the notice of our participants, who only needed to examine about 10 lineups each (around 200 panels). 
An example of one of the problematic lineups is shown in @fig-lineup-problems: many participants selected panel 16 because of the missing ellipse; not a wrong choice, but certainly not the effect we intended to test.

@@

One reason why it is so difficult to generate sampling models for visual explorations is that our visual systems are very, very good at finding differences and unexpected results. 
We re-ran the experiment using a different clustering method, and found that instead of noticing the number of ellipses, when that variable was removed participants instead cued in on the differences in size and shape of the ellipses formed using k-means clustering after the data generating procedure. 
That is, participants could still detect the artificial nature of the induced clusters using other features. 
While it can be difficult to get the data generating method right, it is essential to conducting visual experiments that generalize well beyond the effects shown in a single data set or phenomenon. 
The time and effort invested in this step at the outset of the experiment pays dividends when it allows for clear generalization of the experimental results to an entire statistical concept rather than a single data set.
